---
title: "GSE Dataset Process"
author: "Lin Tian"
date: "2/3/2017"
output: html_document
---

## Load `Biobase` and `GEOquery` packages.

```{r}
library(Biobase)
library(GEOquery)
```

### Download the data by GEO ID

Here, we use `GSE2990` as example. Download the data use `getGEO` function.

```{r}
GSE2990 <- getGEO(GEO="GSE2990", destdir=getwd())
class(GSE2990)
```

The class of `GSE2990` is **list**. To convert it to `Biobase` class, you can use the following code.

```{r}
eset <- GSE2990[[1]]
class(eset)

## Sample Gene Expression/Profiling Data
View(exprs(eset))

## Sample Annotation/Clinical Data
View(pData(eset))
```

You can check whether the dataset has been normalized using boxplot.

```{r}
dat <- exprs(eset)
dim(dat)
boxplot(dat[, 1:20]) ## plot first 20 samples
```

* If there is very big range between minumum and maximum value, you can apply `log2` transformation to the data set `dat <- log2(dat)`.
* If the median intensity is not in the same level, you can use **median normalization** or **loess normalization**.
* If you want to compare across different datasets, you can use **quantile normalization**.

Here are the code for normalization:

```{r}
dat.norm.med <- normalizeMedianAbsValues(dat)
boxplot(dat.norm.med, main="Median normalization")

dat.norm.loess <- normalizeCyclicLoess(dat)
boxplot(dat.norm.loess, main="Loess normalization")

dat.norm.quantile <- normalizeBetweenArrays(dat, method="quantile")
boxplot(dat.norm.quantile, main="Quantile normalization")
```

## Collapse probe name to gene names using Biomart

```{r}
library(biomaRt)
ensembl <- useMart("ENSEMBL_MART_ENSEMBL", host="www.ensembl.org")
listDatasets(ensembl) ## list available data set (species)
ensembl <- useDataset("hsapiens_gene_ensembl",mart=ensembl)
listFilters(ensembl) ## list available filters (probe set/microarray platform)
map <- getBM(mart=ensembl, attributes=c("affy_hg_u133a", "hgnc_symbol"), filters="affy_hg_u133a", values=rownames(dat))
map <- map[map$hgnc_symbol!="", ]
sample.probe <- rownames(dat)[rownames(dat) %in% map$affy_hg_u133a]
ambiguous <- sapply(sample.probe, function(x) length(unique(map$hgnc_symbol[map$affy_hg_u133a==x]))>1) ## remove probes that map to multiple genes
sample.probe <- sample.probe[!ambiguous]
map <- map[map$affy_hg_u133a %in% sample.probe, ]
genenames <- unique(map$hgnc_symbol[map$affy_hg_u133a %in% sample.probe])

dat.gene <- matrix(NA, nrow=length(genenames), ncol=ncol(dat))
rownames(dat.gene) <- genenames
colnames(dat.gene) <- colnames(dat)
for(g in genenames) {
  probes <- as.character(map$affy_hg_u133a[map$hgnc_symbol==g])
  if(length(probes)==1) {
    dat.gene[g, ] <- dat[probes, ]
  }
  else{
    dat.gene[g, ] <- apply(dat[probes, ], 2, max)
  }
}
```

