---
title: "Source Code for Figure 1"
date: "2/1/2016"
---

## Load METABRIC Discovery dataset and essential packages

```{r}
setwd("~/vesselNormalization")
load("~/rstudio/lintian/metabric.exp.RData") # load METABRIC Discovery Dataset; you need apply to the access to the dataset
load("~/vesselNormalization/data/GO0001525_Angiogenesis.RData") # load 377 genes in angiogenesis pathway (GO:0001525)
library(limma) ## package for gene differential expression analysis
```

The gene expression information is stored in **metabric.d.exp**; row: patient/sample ids; col: gene names
The patient annotation information is stored in **metabric.d.ann**; row: patient/sample ids; col: clinical parameters

## Select Patients

Only patients who were survived or died of cancer are kept.
Patients whose survival information is lost or were not died of cancer are removed.

The patients who were alived were relabeled as 'a'; the patients who died of cancer were relabeled with 'dds' (disease specific death).

```{r}
selected_patient.d <- which(metabric.d.ann$last_follow_up_status == "a" | metabric.d.ann$last_follow_up_status == "d-d.s.")
selected_patient.d.exp <- metabric.d.exp[selected_patient.d, angiogenesis]
selected_patient.d.ann <- metabric.d.ann[selected_patient.d, ]
selected_patient.d.ann$last_follow_up_status <- ifelse(selected_patient.d.ann$last_follow_up_status=="d-d.s.", "dds", "a")
```

## Differential Genes Selection

The output is **Supplementary Table 1a, b**.

```{r}
survival <- factor(selected_patient.d.ann$last_follow_up_status)
design <- model.matrix(~0+survival) # Design matrix
contrast.matrix <- makeContrasts(contrasts="survivala - survivaldds", levels=design) # Make contrasts
fit <- lmFit(t(selected_patient.d.exp), design) # Linear model fitting
fit1 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit1) # Bayes test
dif <- topTable(fit2, coef="survivala - survivaldds", n=nrow(fit2), adjust.method="fdr") # calculate the fold change, p value and false discovery rate (fdr)
dif <- dif[dif[, "adj.P.Val"]<0.01, ] # keep genes whose false discovery rate is below 0.01

# Get GPAGs (up-regulated in patients whose survival status is 'a': survivala - survivaldds > 0)
GPAGs <- subset(dif, logFC > 0)
GPAGs <- GPAGs[order(GPAGs$logFC, decreasing = TRUE), ]
GPAGs

# Get PPAGs (down-regulated in patients whose survival status is 'a': survivala - survivaldds < 0)
PPAGs <- subset(dif, logFC < 0)
PPAGs <- PPAGs[order(PPAGs$logFC, decreasing = FALSE), ]
PPAGs
```

## Visualization of GPAGs and PPAGs Using Heatmap (Fig. 1a)

```{r}
library(gplots) ## For plot the heatmap
library(amap) ## For calculate the distance
library(dendextend) ## For draw the dendrogram
library(RColorBrewer)
select.ang.genes <- c(rownames(GPAGs), rownames(PPAGs)) ## combine GPAG names and PPAG names
## Row Cluster Labeling
row_dist <- Dist(t(selected_patient.d.exp[, select.ang.genes]), method="spearman")
rdend <- as.dendrogram(hclust(row_dist, method="ward.D"))
rowv <- rdend %>% set("branches_lwd", 2) %>%  set("by_labels_branches_lty", value=rownames(PPAGs), TF_values = c(3,Inf))
## Column Cluster Labeling
col_dist <- Dist(selected_patient.d.exp[, select.ang.genes], method="kendall")
colv <- color_branches(as.dendrogram(hclust(col_dist, method="ward.D")), k=2)

hmcol <- c(rep("#B35806", 90), colorRampPalette(brewer.pal(11, "PuOr"))(50), rep("#542788", 100))

heatmap.2(t(selected_patient.d.exp[, select.ang.genes]), col=hmcol, Rowv=rowv, Colv=colv, trace = "none", scale ="row", labRow=NA, labCol=NA, keysize = 1.2, key.title="Color Key")
```

## Survival Analysis on Two Clusters in the Heatmap (Fig. 1b)

```{r}
library(survival)
patient.label <- cutree(hclust(col_dist, method="ward.D"), k=2) ## Get labeling information of the two clusters in the heatmap
survdiff(Surv(selected_patient.d.ann$T/30, selected_patient.d.ann$DSS) ~ patient.label) # p = 4.47e-09
coxph(Surv(selected_patient.d.ann$T/30, selected_patient.d.ann$DSS) ~ patient.label) # Hazard Ratio > 2
plot(survfit(Surv(selected_patient.d.ann$T/30, selected_patient.d.ann$DSS) ~ patient.label), col=c("#009681", "#CC476B"), lty=1, lwd=1, xlab = "Months", ylab = "Disease-Free Survival Prob.", ylim=c(0, 1))
legend("bottomleft", c("Cluster I (151/333)", "Cluster II (109/475)"), col=c("#CC476B", "#009681"), lty=1, bty = "n", lwd=1)
```


## Correlation of GPAGs and PPAGs with T Cell Activation (Fig. 1f)

To test the correlation of GPAGs, PPAGs with T cell activation, we used the samples in the combined METABRIC dataset (Discovery and Validation).

```{r}
library(LSD)
load("~/vesselNormalization/data/GO0042110_TcellActivation.RData")

## Correlation between GPAGs and TCR activation
cor.test(apply(metabric.a.exp[, rownames(GPAGs)], 1, sum), apply(metabric.a.exp[, GO0042110], 1, sum))

## Correlation between PPAGs and TCR activation
cor.test(apply(metabric.a.exp[, rownames(PPAGs)], 1, sum), apply(metabric.a.exp[, GO0042110], 1, sum))

TCR_sig <- rbind(data.frame(ang=apply(metabric.a.exp[, rownames(GPAGs)], 1, sum), TCR=apply(metabric.a.exp[, GO0042110], 1, sum), type="GPAGs"), data.frame(ang=apply(metabric.a.exp[, rownames(PPAGs)], 1, sum), TCR=apply(metabric.a.exp[, GO0042110], 1, sum), type="PPAGs"))
TCR_sig$norm_ang <- scale(TCR_sig$ang)
TCR_sig$norm_TCR <- scale(TCR_sig$TCR)

heatscatter(as.numeric(TCR_sig[TCR_sig$type=="GPAGs", "norm_ang"]), as.numeric(TCR_sig[TCR_sig$type=="GPAGs", "norm_TCR"]), colpal="bl2gr2rd", xlab="Normalized Angiogenesis Signature", ylab="Normalized TCR signaling", cor=TRUE, cex.main=1, main="Correlation of GPAGs with T Cell Activation;")

heatscatter(as.numeric(TCR_sig[ TCR_sig$type=="PPAGs", "norm_ang"]), as.numeric(TCR_sig[TCR_sig$type=="PPAGs", "norm_TCR"]), colpal="bl2gr2rd", xlab="Normalized Angiogenesis Signature", ylab="Normalized TCR signaling", cor=TRUE, cex.main=1, main="Correlation of PPAGs with T Cell Activation;")
```

